Is it Better to be a Moose or a Black Bear?

### 2024-04-21 paper/paper.qmd
Finished the intro of the quarto document and started the discussion section

---
title: "Is it Better to be a Moose or a Black Bear?"
subtitle: "An Analysis of Ontario Wildlife Harvest and Active Hunter Data From 2006 to 2023"
author: Samantha Barfoot
thanks: "Code and data are available at: https://github.com/SamanthaBarfoot/wildlife_hunting.git"
date: today
date-format: long
abstract: "This paper analyzes Ontario wildlife harvest counts, specifically moose and black bears, and the number of active hunters based on region. The analysis examines harvest counts from 2012 to 2023 across northern, southeastern, and southwestern Ontario. It was found that..."
format: pdf
number-sections: true
table-of-contents: true
bibliography: references.bib
---

@@

Therefore, it is important to properly regulate and monitor the wildlife populations as well as the hunting and harvest numbers so that species are not put at risk by over hunting. Additionally, it is important to analyze how these levels change over the years and what might cause these changes. Alongside over hunting, having too large a population of a certain species can also put a strain on the ecosystem in which case hunting is an important method of regulating population. This paper explores the trends of active hunter and harvest numbers from 2006-2023 for moose and 2021-2023 for black bears.

WHAT WAS FOUND

The estimand of this paper is the correlation between the number of active hunters in Ontario and the number of annual moose and black bear harvests. Specifically, we examine how these numbers vary with regard to region and time

@@

MORE ABOUT THE SECTIONS

@@

```{r}
#| echo: false
#| warning: false
#| label: fig-harvestmoosetypes
#| fig-cap: "Number of cow, bull, and calf moose types harvested between 2012 and 2023 in the northern (A), southeastern (B), and southwestern (C) region"

@@

TABLE?

@@

## Total Harvests

@@

## Active Hunters

@@

@@

By examining the number of hunters against the number of harvests for both moose (@fig-harvesthuntermoose) and black bears (@fig-harvesthuntersbear) we can notice a few trends. Overall, for both moose and black bears in all three regions, as the number of hunters increases as does the total harvest. However, there are still many fluctuations in this trend, especially in the southeastern region. For both moose and black bear there are numerous dips and peaks where, as the number of hunters increase, the total harvest is significantly less or more than the overall average.

@@

plot1 <- 
  ggplot()+
  geom_line(data = north_moose, aes(x = `Active Hunters`, y = `Total Harvest`), color = "red") +
  theme_minimal()+
  labs(tag = "A")

plot2 <-
  ggplot()+
  geom_line(data = se_moose, aes(x = `Active Hunters`, y = `Total Harvest`), color = "red") +
  theme_minimal() +
  labs(tag = "B")

plot3 <-
  ggplot()+
  geom_line(data = sw_moose, aes(x = `Active Hunters`, y = `Total Harvest`), color = "red") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 70))
  labs(tag = "C")

@@

plot1 <- 
  ggplot()+
  geom_line(data = north_bear, aes(x = `Active Hunters`, y = Harvest), color = "blue") +
  theme_minimal()+
  labs(tag = "A")

plot2 <-
  ggplot()+
  geom_line(data = se_bear, aes(x = `Active Hunters`, y = Harvest), color = "blue") +
  theme_minimal() +
  labs(tag = "B")

plot3 <-
  ggplot()+
  geom_line(data = sw_bear, aes(x = `Active Hunters`, y = Harvest), color = "blue") +
  theme_minimal() +
  labs(tag = "C")

@@

@@

## What was done

## Something we learned about the world

## Another thing we learned about the world

@@

>>>>

---
title: "Is it Better to be a Moose or a Black Bear?"
subtitle: "An Analysis of Ontario Wildlife Harvest and Active Hunter Data From 2006 to 2023"
author: Samantha Barfoot
thanks: "Code and data are available at: https://github.com/SamanthaBarfoot/wildlife_hunting.git"
date: today
date-format: long
abstract: "This paper analyzes Ontario wildlife harvest counts, specifically moose and black bears, and the number of active hunters based on region. The analysis examines harvest counts from 2012 to 2023 across northern, southeastern, and southwestern Ontario. It was found that moose harvests have decreased over the past 20 years whereas black bear hunting has stayed relativly consistant. The implications and causiontions of these trends are important for species population in Ontario as well as the Ontario hunting industry."
format: pdf
number-sections: true
table-of-contents: true
bibliography: references.bib
---

@@

Therefore, it is important to properly regulate and monitor the wildlife populations as well as the hunting and harvest numbers so that species are not put at risk by over hunting. Additionally, it is important to analyze how these levels change over the years and what might cause these changes. Alongside over hunting, having too large a population of a certain species can also put a strain on the ecosystem in which case hunting is an important method of regulating population. The Ontario hunting industry is a significant economic contributor and, therefore, it is important to the province for hunting practices to continue. This paper examines these hunting activities in Ontario and how they have changed over the past few years. Specifically focusing on two majorly hunted species, moose and black bear, which are favoured by trophy hunters. Using data from the Ontario Data Catalogue active hunter and harvest numbers from 2006-2023 for moose and 2021-2023 for black bears are analyzed. From this analysis, it was found that moose harvests have drastically decreased since 2006 as has the number of active moose hunters. In comparison, black bear harvesting and hunting has stayed relatively consistent. While moose hunting has decreased there is still a significant population of moose hunters, especially compared to black bear hunters.

The estimand of this paper is the correlation between the number of active hunters in Ontario and the number of annual moose and black bear harvests. Specifically, we examine how these numbers vary with regard to region and time.

@@

@@

```{r}
#| echo: false
#| warning: false
#| label: fig-harvestmoosetypes
#| fig-cap: "Total number of cow, bull, and calf moose types harvested between 2012 and 2023"

@@

@@

## Total Harvest Trends

@@

By examining the data sets we also learn that for black bears the average harvest between 2012 and 2023 is 3,160 in the northern region, 2,046 in the southeastern region, and 177 in the southwestern region. In compression, for moose between 2006 and 2023 it is 3,356 in the northern region, 1,072 in the southeastern region, and 129 in the southwestern region. The maximum number of moose harvests was in 2006 with 7,342 total harvests.The minimum number of moose harvests was in 2023 with 2,915 total harvests. Whereas, for black bears the maximum number of harvests was in 2016 with 8,152 harvests. The minimum was in 2021 with 3,785 active hunters.

## Active Hunter Trends

@@

By examining the data sets we also learn that for black bears the average number of active hunters between 2012 and 2023 is 12,736 in the northern region, 12,717 in the southeastern region, and 499 in the southwestern region. In compression, for moose between 2006 and 2023 it is 49,666 in the northern region, 19,298 in the southeastern region, and 1332 in the southwestern region. The maximum number of active moose hunters was in 2010 with 88,369 total hunters. The minimum number of active hunters was in 2023 with 36,069 total hunters. Whereas, for black bears the maximum number of hunters was in 2016 with 31,480 hunters. The minimum was in 2013 with 20,891 active hunters.

@@

By examining the number of hunters against the number of harvests for both moose (@fig-harvesthuntermoose) and black bears (@fig-harvesthuntersbear) we can notice a few trends. Overall, for both moose and black bears in all three regions, there is a linear relationship between the number of hunters and the number of harvests. From the grpahs we can see that as the number of hunters increases as does the total harvest. However, there are still many fluctuations in this trend, especially in the southeastern region. For both moose and black bear there are numerous dips and peaks where, as the number of hunters increase, the total harvest is significantly less or more than the overall average.

@@

plot1 <- 
  ggplot()+
  geom_line(data = north_moose, aes(x = `Active Hunters`, y = `Total Harvest`), color = "red") +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 70))+
  labs(tag = "A")

plot2 <-
  ggplot()+
  geom_line(data = se_moose, aes(x = `Active Hunters`, y = `Total Harvest`), color = "red") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 70))+
  labs(tag = "B")

plot3 <-
  ggplot()+
  geom_line(data = sw_moose, aes(x = `Active Hunters`, y = `Total Harvest`), color = "red") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 70))+
  labs(tag = "C")

@@

plot1 <- 
  ggplot()+
  geom_line(data = north_bear, aes(x = `Active Hunters`, y = Harvest), color = "blue") +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 70))+
  labs(tag = "A")

plot2 <-
  ggplot()+
  geom_line(data = se_bear, aes(x = `Active Hunters`, y = Harvest), color = "blue") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 70))+
  labs(tag = "B")

plot3 <-
  ggplot()+
  geom_line(data = sw_bear, aes(x = `Active Hunters`, y = Harvest), color = "blue") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 70))+
  labs(tag = "C")

@@

Overall, from all of these graphs we can see that every year less and less people are actively hunting moose. The number of people hunting moose and the number of moose being harvested is decreasing. Whereas, hunting black bears peaked greatly in 2016 but has since gone back to its pre 2016 harvest levels and appears to be staying relatively consistent. Additionally, we can see that while there are often more moose hunters than black bear hunters, more black bears are harvested yearly then moose on average.

@@

This paper has analyzed the correlation between the number of active hunters and the number of moose and black bear harvests in Ontario and the impact of time and region on these numbers. Additionally it examined the number of harvests against the number of active hunters. The analysis focused on three main regions, the northern, southeastern, and southwestern areas of Ontario. This was done with the analysis of two data sets obtained from the Ontario Data Catalogue.

## Ontario moose hunting is declining

## Black bears are popular hunting targets in Ontario

@@

The primary weakness of this analysis is its lack of depth. This is because it only focuses on and uses data with two main variables, number of active hunters and harvests. From the data alone it is difficult to know the exact reasons for certain trends, such as the decrease in moose harvesting without external research. This analysis could have been made stronger if paired with moose and black bear population data. More types of data and variables would have added more depth to this analysis.

Another weakness is that the data doesn't cover a very large time span. The black bear data is only from 2012 to 2023 whereas the moose data covers a larger time frame from 2006 to 2023. Analyzing a data set with a larger range would provide more depth. This would also be beneficial as trends related to populations provide more insight when analyzed over a large period of time. The moose data only shows decreasing harvest rates so it would be interesting to see when they began increasing.
